---
title: "Homework 4"
author: Denis Ostroushko
format: 
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
editor: source
execute:
  echo: false
  warning: false
  message: false 
---

<!-- relative risk = hazard ratio --> 

```{r}

library(survival)
library(km.ci)
library(KMsurv)
library(tidyverse)
library(kableExtra)
library(survminer)
library(FHtest)
library(data.table)
library(survRM2)

```

# Problem 1

### (a)

Proportional Hazards Model: $\large h(t|Z) = h_0(t) exp(\boldsymbol{Z' \hat \beta})$, 

where $\large \boldsymbol{Z' \hat \beta}$ is a linear combination of predictors. In our case, $\boldsymbol{\hat \beta}$ is a vector of 
length one, a single model estimate that governs estimated differences between groups. 

So, the full expression for the model is: $\large h(t|Z) = h_0(t) exp(\hat \beta * I(Group \ 2))$, thus $Group \ 1$ is a reference level. 

### (b)

We create a dummy variable to conduct regression analysis, set indicator to 0 when an individual is in an 'Untreated' group, and 
1 when in 'Radiated' group

To write down the partial likelihood for the model, we need to specify the hazard function for the survival times. 
For Cox proportional hazards model we assume that the hazard function for each group is proportional to the hazard function for the other group, with a constant hazard ratio over the entire observed timeline. 

Generally, partial likelihood for the Cox model is given by:

L(β) = ∏i[exp(βZi)/(∑j[exp(βZj)I(Tj > Ti)])]^(δi)



$$\large L = \prod_{i=1}^{n} \frac{exp(\beta Z_i)}{\sum_{j\in R(t_i)}exp(\beta Z_j)}$$


where: 

1. $N$ is the sample size
2. $\beta$ is the log hazard ratio for the radiated group compared to the untreated group
3. $X_i$ is the value of the dummy variable for the ith individual (i.e., 0 for untreated, 1 for radiated)
4. $T_i$ is the observed survival time for the ith individual, 
5. $\delta_i$ is an indicator variable that takes the value 1 if the survival time is uncensored 

For this case with one dummy predictor the likelihood function has a simpler form

$$\large L = \prod_{i=1}^{12} \frac{exp(\beta Z_i)}{\sum_{j\in R(t_i)}exp(\beta Z_j)}$$

### (c)

The data patterns and order of events and corresponding groups remains the same, so the likelihood function will remain almost the same. 

### (d)

The baseline hazard rate for the two models will be different due to different time values that go into the hazard equation. 
However, variance and betas should be the same due to similar pattern and ordering of events and corresponding groups. 

# Problem 2

```{r }
data('tongue')

# type: Tumor DNA profile (1=Aneuploid Tumor, 2=Diploid Tumor)

# time: Time to death or on-study time, weeks

# delta: Death indicator (0=alive, 1=dead)

```

### (a)

Summary of fitted Cox PH model is given below. We will use the test statistics and corresponding p-values to conduct the tests we are required. 

```{R}
#| echo: true

tongue$descr <- 
  factor(
    ifelse(tongue$type == 1, "Aneuploid", "Diplod"), 
    levels = c("Diplod", "Aneuploid")
    )

fit <- coxph(Surv(time, delta) ~ descr, method="breslow", data=tongue)
summary(fit)

```
**Score test**

1. Null hypothesis: no difference in survival rates between groups 

2. Alternative hypothesis: some difference in survival rates between groups 

3. Test statistic: 2.75; P-value: 0.1

4. Conclusion: we do not reject null hypothesis due to lack of overwhelming statistical evidence. Although, these results might be suggestive of the fact that there is some difference between the groups 

### (b)

95% confidence interval is given below, we exponentiate model estimate 

```{r}
#| echo: true
exp(confint(fit))
```


### (c)

Keep model with Breslow ties handling. Grab likelihood ratio test statistic and p-value. 

1. Null hypothesis: no difference in survival rates between groups 

2. Alternative hypothesis: some difference in survival rates between groups 

3. Test statistic: 2.61; P-value = 0.1

4. Conclusion: we do not reject null hypothesis due to lack of overwhelming statistical evidence. Although, these results might be suggestive of the fact that there is some difference between the groups 

### (d)

Keep model with Breslow ties handling. Grab Wald test statistic and p-value. 


1. Null hypothesis: no difference in survival rates between groups 

2. Alternative hypothesis: some difference in survival rates between groups 

3. Test statistic: 2.7; P-value = 0.1

4. Conclusion: we do not reject null hypothesis due to lack of overwhelming statistical evidence. Although, these results might be suggestive of the fact that there is some difference between the groups 

### (e)

Breslow

```{r}
r <- coef(summary(fit))[1:3]

names(r) <- c('coef',  'exp(coef)',   'se(coef)' )

r
```

```{r}
fit4 <- coxph(Surv(time, delta) ~ descr, data=tongue, method = "efron")

r <- coef(summary(fit4))[1:3]

names(r) <- c('coef',  'exp(coef)',   'se(coef)' )

r
```

Estimates are almost identical. I speculate that the small number of ties is the reason for a minor discrepancy. 

# Problem 3

```{r}
data('bnct')

bnct$Z1 <- ifelse(bnct$trt == 2, 1, 0)
bnct$Z2 <- ifelse(bnct$trt == 3, 1, 0)

```

### (a)

columns `coef` and `se(coef)` provide beta estimates for Z1 and Z2

```{r}
#| echo: true

fit1 <- coxph(Surv(time, death) ~Z1 + Z2, data=bnct, method = "breslow")

summary(fit1)$coefficient 

```


### (b)

there is no intercept in a cox ph model, so to compare radiated to untreated we need to use estimate and variance for 
radiated group. First, grab a model estimate

```{r}
coef(fit1)[1]
```

Then grab variance of the estimate to get a standard error
```{r}
vcov(fit1)[1,1]
```

Create a confidence interval and exponentiate the end points to get a confidence interval for the comparison 

```{r}
#| echo: true

  exp(
    c(
      coef(fit1)[1] - 1.96 * sqrt(vcov(fit1)[1,1]),
      coef(fit1)[1] + 1.96 * sqrt(vcov(fit1)[1,1])
    )
  )

```

### (c)

To conduct this test we compare the two Cox PH models: with and without two radiation treatment variables. 
Test output is given below: 

```{r}
model_reduced <- coxph(Surv(time, death) ~ 1, data = bnct, method = "breslow")

# Perform the likelihood ratio test using the anova() function
anova(model_reduced, fit1, test = "LRT")
```

**LRT**

1. Null hypothesis: $\beta_{Z_1} = \beta_{Z_2} = 0$

2. Not all coefficients are equal to 0 

3. Test statistic: 27.371; p-value < 0.001

4. Conclusion: there is strong evidence that not all beta estimates are equal to 0 and some radiation treatment has effect on survival rates

**Wald Test**

Code for the test is given below: 

```{r}
#| echo: true
library(aod)
wald_test <- wald.test(b = coef(fit1), Sigma = vcov(fit1), Terms = 2:length(coef(fit1)))

# View the results
wald_test
```


1. Null hypothesis: $\beta_{Z_1} = \beta_{Z_2} = 0$

2. Not all coefficients are equal to 0 

3. Test statistic: 22.0; p-value < 0.001

4. Conclusion: there is strong evidence that not all beta estimates are equal to 0 and some radiation treatment has effect on survival rates


### (d)

First, obtain the difference between coefficients: 

```{r}
#| echo: true
diff = coef(fit1)[1] - coef(fit1)[2]
diff
```

Now obtain pooled variance, $\large Var(Diff) = Var(\beta_1) + Var(\beta_2) - 2 Cov(\beta_1, \beta_2)$

```{r}
#| echo: true
#| 
pooled_var = vcov(fit1)[1,1] + vcov(fit1)[2,2] - 2*vcov(fit1)[1,2]
pooled_var
```

Conduct Z-test

```{R}
#| echo: true

(diff - 0)/ sqrt(pooled_var)

```

1. Null hypothesis: $\beta_1 = \beta_2 = 0$

2. Alternative hypothesis: $\beta_1 \neq \beta_2$

3. Test statistic: `r (diff - 0)/ sqrt(pooled_var)`; P-value: `r 2 * pnorm( abs((diff - 0)/ sqrt(pooled_var)), lower.tail=FALSE)`

4. Conclusion: we have strong evidence to believe that the two coefficients are different. Therefore, effect of two treatments on the 
survival rates are not the same 

### (e)

Using previous part, obtain a confidence interval for the difference using quantiles of normal distribution. Then exponentiate the 
endpoints to get a confidence interval for the relative risk. 

```{r}

exp(
  c(
    diff - 1.96 * sqrt(pooled_var),
    diff + 1.96 * sqrt(pooled_var)
  )
)

```

### (e) again 

same as part (c)

```{r}
anova(model_reduced, fit1, test = "Chisq")
```

Same conclusion follows 

### (g)

same as part (c)

```{r}
wald_test <- 
  wald.test(b = coef(fit1)[c("Z1", "Z2")], 
            Sigma = vcov(fit1)[c("Z1", "Z2"), c("Z1", "Z2")], 
            Terms = 2:length(coef(fit1)))

# View the results
wald_test
```

Same conclusion follows 