---
title: "Homework 4"
author: Denis Ostroushko
format: 
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
editor: source
execute:
  echo: false
  warning: false
  message: false 
---

```{r}

library(survival)
library(km.ci)
library(KMsurv)
library(tidyverse)
library(kableExtra)
library(survminer)
library(FHtest)
library(data.table)
library(survRM2)

```

# Problem 1

### (a)

Proportional Hazards Model: $\large h(t|Z) = h_0(t) exp(\boldsymbol{Z' \hat \beta})$, 

where $\large \boldsymbol{Z' \hat \beta}$ is a linear combination of predictors. In our case, $\boldsymbol{\hat \beta}$ is a vector of 
length one, a single model estimate that governs estimated differences between groups. 

So, the full expression for the model is: $\large h(t|Z) = h_0(t) exp(\hat \beta * I(Group \ 2))$, thus $Group \ 1$ is a reference level. 

### (b)

We create a dummy variable to conduct regression analysis, set indicator to 0 when an individual is in an 'Untreated' group, and 
1 when in 'Radiated' group

To write down the partial likelihood for the model, we need to specify the hazard function for the survival times. 
For Cox proportional hazards model we assume that the hazard function for each group is proportional to the hazard function for the other group, with a constant hazard ratio over the entire observed timeline. 

Generally, partial likelihood for the Cox model is given by:

$$\large L = \prod_{i = 1} ^{N} h_0(T_i)^{\delta_i} * [exp(\boldsymbol{\beta ' X_i})]^{\delta_i} * exp(-H_0(T_i) \ exp(\boldsymbol{\beta'X_i}))$$


where: 

1. N is the sample size, 
2. $\beta$ is the log hazard ratio for the radiated group compared to the untreated group
3. $X_i$ is the value of the dummy variable for the ith individual (i.e., 0 for untreated, 1 for radiated)
4. $T_i$ is the observed survival time for the ith individual, 
5. $\delta_i$ is an indicator variable that takes the value 1 if the survival time is uncensored 


### (c)

### (d)

# Problem 2

```{r }
data('tongue')

# type: Tumor DNA profile (1=Aneuploid Tumor, 2=Diploid Tumor)

# time: Time to death or on-study time, weeks

# delta: Death indicator (0=alive, 1=dead)

```

### (a)

**Breslow** 

```{R}
#| echo: true

fit <- coxph(Surv(time, delta) ~ type, method="breslow", data=tongue)
summary(fit)

```
1. Null hypothesis: no difference in survival rates between groups 

2. Alternative hypothesis: some difference in survival rates between groups 

3. P-value: `r round(coef(summary(fit))[5], 4)`

4. Conclusion: we do not reject null hypothesis due to lack of overwhelming statistical evidence. Although, these results might be suggestive of the fact that there is some difference between the groups 

**Score test**

```{r}
#| eho: true

score <- summary(fit)$sctest[1]
score_p_val <- summary(fit)$sctest[3]
```

1. Null hypothesis: no difference in survival rates between groups 

2. Alternative hypothesis: some difference in survival rates between groups 

3. P-value: `r round(score_p_val, 4)`

4. Conclusion: we do not reject null hypothesis due to lack of overwhelming statistical evidence. Although, these results might be suggestive of the fact that there is some difference between the groups 

### (b)

95% confidence interval is given below

```{r}
#| echo: true
exp(confint(fit))
```


### (c)

```{r}
#| echo: true

fit2 <- coxph(Surv(time, delta) ~ type, method="exact", data=tongue)
summary(fit2)

```

P-value: `r round(coef(summary(fit2))[5], 4)`

### (d)

Wald is just regular model without special method specification 

```{r}
fit3 <- coxph(Surv(time, delta) ~ type, data=tongue)
summary(fit3)
```

P-value: `r round(coef(summary(fit3))[5], 4)`

### (e)

Breslow

```{r}
r <- coef(summary(fit))[1:3]

names(r) <- c('coef',  'exp(coef)',   'se(coef)' )

r
```

```{r}
fit4 <- coxph(Surv(time, delta) ~ type, data=tongue, method = "efron")

r <- coef(summary(fit4))[1:3]

names(r) <- c('coef',  'exp(coef)',   'se(coef)' )

r
```

# Problem 3

### (a)

### (b)

### (c)

### (d)

### (e)

### (g)

