---
title: "Homework 2"
author: "Denis Ostroushko"
format: 
  pdf:
    toc: false 
    number-sections: false
    colorlinks: true
editor: source
execute: 
  eval: true
  echo: false
  warning: false
  message: false 
---

```{r}

library(survival)
library(tidyverse)
library(survminer)
library(muhaz)
library(km.ci)
library(KMsurv)

```

# Problem 1

### (i)

We are not able to calculate variance with the information given in this problem. We need the number of 
events and number of observations at risk for each time point leading up to the time point $j$. 

### (ii)

For an asymptotically normal random variable, either the delta method or an endpoint transformation 
produce desired outcome. 

We will use the endpoint transformation method to find a confidence interval for the log-odds of surviving 
past a time point $t_j$. 

First, obtain a confidence interval for the probability of surviving, which is given by 

$$\large \widehat{S(t_j)} \pm 1.96 * \sqrt{\widehat{Vat(S(t_j))}} = (lb, up)$$
The, for the log-odds of survival, a confidence interval is given by 

$$\Large (log \frac{lb}{1-lb}, log \frac{ub}{1-ub})$$
**Delta Method** 

if we have to determine the variance, as stated in the Note section, we would use the delta method approach
to get variance of $T(S(t_j))$. 

First, define $\large g(S(t_j)) = log \frac{S(t_j)}{1-S(t_j)}$

Then, $\large g'(S(t_j))^2 = \frac{d}{dS} g(S(t_j)) = (\frac{1}{S(t_j)} + \frac{1}{1 - S(t_j)})^2$

Therefore, $\large Var[T(S(t_j))] = (\frac{1}{S(t_j)} + \frac{1}{1 - S(t_j)})^2 * Var(S(t_j))$ 

We will need to compute this variance at $t_j$ using the value of $S(t_j)$ and its variance. 

Then we compute a 95% confidence interval: $\large T(S(t_j)) \pm 1.96 * \sqrt{ Var[T(S(t_j))]}$

### (iii)

Alternative hypothesis: $H_a: S_1(t_j) \neq S_0(t_j)$

Test statistic: $\frac{( S_1(t_j) -  S_0(t_j)) - 0}{\sqrt{Var( S_1(t_j)) + Var( S_0(t_j))}}$

Since both $S_1(t_j)$ and $S_0(t_j)$ are normally distributed, we can conduct the two sided test using 1.96 - 
a 97.5th quantile of standard normal distribution. 

If the absolute value of our test statistic is greater than 1.96, then we can reject the null hypothesis and 
conclude that the survival probability past time point $t_j$ is different for the two groups. 


### (iv)

$$\Large ??$$ 

# Problem 2

```{r}
data(kidney)

# head(kidney)
```

### (i)

@fig-km-prob2 shows the two KM survival curves for the two groups. It appears that group 2, with a 
percutaneously placed catheter, has more evidence that this method of placement delays time of infection 
more effectively. 

```{r}
#| label: fig-km-prob2
#| fig-cap: Kaplan Meier Estiamtes for time to infection in weeks the two groups  
#| fig-width: 6
#| fig-height: 6

fit <- survfit(Surv(time, delta) ~type ,data = kidney)

ggsurvplot(fit, data = kidney, 
           risk.table = F, 
           pval = TRUE,
           palette = c("#E41A1C", "#377EB8"), 
           title = "Survival Estimates by Group",
           conf.int = TRUE)

```

### (ii)

<!-- https://meichenlu.com/2019-03-03-Survival-Analysis-3-Implementation/ --> 

```{r}

kphaz.res <- kphaz.fit(kidney$time, kidney$delta, method="nelson")

kphaz.res %>% data.frame() %>% 
  mutate(cumhaz = cumsum(haz)) -> CH_estiamte_date 

ggplot(data = CH_estiamte_date, 
       aes(x = time, 
           y = cumhaz)) + geom_step()

# hazard at time 5 
CH_estiamte_date %>% filter(time == 5) %>% select(haz) %>% unlist()


```

### (iii)

```{r}

fit_all <- survfit(Surv(time, delta) ~1 ,data = kidney)

s.km.fit = summary(fit_all, rmean=36)

s.km.fit$table["rmean"]


```

### (iv)

```{r}
# group 1 

fit_type1 <- survfit(Surv(time, delta) ~1 ,data = kidney %>% filter(type == 1))
s.km.fit_type1 = summary(fit_type1, rmean=36)

s.km.fit_type1$table["rmean"]
s.km.fit_type1$table["se(rmean)"]

```

```{r}
# group 2 
fit_type2 <- survfit(Surv(time, delta) ~1 ,data = kidney %>% filter(type == 2))
s.km.fit_type2 = summary(fit_type2, rmean=36)

s.km.fit_type2$table["rmean"]
s.km.fit_type2$table["se(rmean)"]

```

### (v)

```{r}
# group 1

HW.plain <- km.ci(fit_type1, conf.level=0.9, method="hall-wellner") 
plot(HW.plain, xlab = "Time", ylab = "Survival Probability", col="red")

```

```{r}
# group 2

HW.plain <- km.ci(fit_type2, conf.level=0.9, method="hall-wellner") 
plot(HW.plain, xlab = "Time", ylab = "Survival Probability", col="red")

```

# Problem 3

```{r }

data(kidtran) 
kidtran_f <- kidtran %>% filter(gender == 2 & race == 1)

```

### (i)

### (ii)

### (iii)

### (iv)

### (v)

### (vi)

### (vii)


