---
title: "Homework 5"
author: Denis Ostroushko
format: 
  pdf:
    toc: false
    number-sections: false
    colorlinks: true
editor: source
execute:
  echo: false
  warning: false
  message: false 
---

```{r}

library(survival)
library(km.ci)
library(KMsurv)
library(tidyverse)
library(kableExtra)
library(survminer)
library(FHtest)
library(data.table)
library(survRM2)

```

# Exercise 1

```{R}
data("larynx")

larynx$stage <- 
  factor(
    larynx$stage, 
    levels = unique(sort(larynx$stage))
  )

larynx$diagyr_strata <- 
  factor(
  with(larynx, 
       case_when(
         diagyr < 75 ~ "Prior to 1975", 
         T ~ "1975 and later"
       ))
)
```

### (a)

R output for a regression model is given below. All three tests suggest that there is evidence to conclude that some variables included in the model are helpful at explaining the outcome. 

```{r}
cox_1_a <- coxph(Surv(time = time, event = delta) ~ age + stage, data = larynx)
summary(cox_1_a)
```

### (b)

Stratification by diagnosis year produces similar results. 
Coefficients, and therefore hazard rations, tell a similar story between the two models. Standard errors appear similar as well. 

```{r}
cox_1_b <- coxph(Surv(time = time, event = delta) ~ age + stage + strata(diagyr_strata), data = larynx)
summary(cox_1_b)
```

### (c)

Code below conducts a likelihood test for the two models.

```{r}
#| echo: true
anova(cox_1_a, cox_1_b, test="Chisq")
```

While we saw that the results of the two models' estimates are similar, it appears that the inclusion of the  stratification variable improves model fit. 

### (d)

I am including code below that I used to conduct the Wald test. Following the book instructions, I am comparing each corresponding estimate between the two strata. 

Walk test suggests that model parameters are similar for the 
two strata. 

```{r}
#| echo: true
coef1 <- summary(cox_1_a)$coefficients
coef2 <- summary(cox_1_b)$coefficients

waldtest_table <- 
  data.frame(
    test_stat = (coef1[,1] - coef2[,1])^2 / (coef1[,3]^2 + coef2[,3]^2)
  )

waldtest_table$p_val = pchisq(waldtest_table$test_stat, lower.tail = F, df = 1)

waldtest_table
```

# Exercise 2

```{r}
data(rats)

# summary(rats$litter)

rats$litter <- as.factor(rats$litter)
```

### (a)

Exponentiated coefficient and end points of the confidence interval produce desired output for the non-stratified cox regression model 

```{r}
#| echo: false
cox_2_a <- coxph(Surv(time = time, event = tumor) ~ trt, data = rats)

exp(summary(cox_2_a)$coefficients[1])

exp(confint(cox_2_a))
```

### (b)

Stratified model produce similar output and results. 

```{r}
cox_2_b <- coxph(Surv(time = time, event = tumor) ~ trt + strata(litter), data = rats)

exp(summary(cox_2_b)$coefficients[1])

exp(confint(cox_2_b))
```

# Exercise 3


```{r}
data(tongue)

tongue$type_descr <- 
  with(tongue, 
       case_when(
         type == 1 ~ 'Aneuploid Tumor', 
         T ~ 'Diploid Tumor'
       ))


```

### (a)


Plot below provides cumulative hazards for the two groups. Visually, they do not appear proportional over time. 

```{r}
#| eval: false

# Calculation of this crap by hand 
cox_3_a <- coxph(Surv(time = time, event= delta) ~ type_descr, data = tongue )

cumhaz_plot <- 
  data.frame(
    cumhaz = 
      c(
        basehaz(coxph(Surv(time = time, event= delta) ~ 1, data = tongue %>% filter(type_descr == 'Aneuploid Tumor')))[,1],
        basehaz(coxph(Surv(time = time, event= delta) ~ 1, data = tongue %>% filter(type_descr != 'Aneuploid Tumor')))[,1]
      ),
    time = 
      c(
        basehaz(coxph(Surv(time = time, event= delta) ~ 1, data = tongue %>% filter(type_descr == 'Aneuploid Tumor')))[,2],
        basehaz(coxph(Surv(time = time, event= delta) ~ 1, data = tongue %>% filter(type_descr != 'Aneuploid Tumor')))[,2]
      ),
     group = 
      c(
        rep('Aneuploid Tumor', 
            length(basehaz(coxph(Surv(time = time, event= delta) ~ 1, data = tongue %>% filter(type_descr == 'Aneuploid Tumor')))[,1])),
        rep('Diploid Tumor', 
            length(basehaz(coxph(Surv(time = time, event= delta) ~ 1, data = tongue %>% filter(type_descr != 'Aneuploid Tumor')))[,1]))
      )
  )

ggplot(data = cumhaz_plot, 
       aes(x = time, y = (cumhaz), color = group)) + geom_step()

```

```{r}
cox_3_a <- coxph(Surv(time = time, event= delta) ~ type_descr, data = tongue )

plot_obj <- survfit(Surv(time = time, event= delta) ~ type_descr, data = tongue )

survminer::ggsurvplot(plot_obj, fun = "cumhaz", legend = "bottom")
```

### (b)

Using some manipulation of the data used for part (a), I carry observed values of cumulative hazard forward and have some data at each time point 1,2,3 ... 

Plot below provides differences of log cumualtive hazards between the two groups. 

```{R}
#| echo: false 
ch_df <- 
  data.frame(
    time = plot_obj[['time']], 
    ch = plot_obj[['cumhaz']]
  )

ch_df$diff = with(ch_df, lag(time) - time)

ch_df$row = as.numeric(rownames(ch_df))
ch_df$group = "1"

ch_df[ch_df$row >= as.numeric(rownames(ch_df[ch_df$diff > 0 & !is.na(ch_df$diff), ])), ]$group = "2"

ch_df <- ch_df %>% select( -diff)

ch_df_2 <- 
  data.frame(
    group = c(
      rep("1", max(ch_df[ch_df$group == '1', ]$time)),
      rep("2", max(ch_df[ch_df$group == '2', ]$time))
    ), 
    time = c(
      seq(from = 1, to =  max(ch_df[ch_df$group == '1', ]$time), by = 1),
      seq(from = 1, to =  max(ch_df[ch_df$group == '2', ]$time), by = 1)
    )
  )

ch_df_2 <- 
  ch_df_2 %>% 
  left_join(ch_df, by = c("time", "group")) %>% 
  arrange(group, time)

ch_df_2 <- ch_df_2 %>% select(-row)

for(i in 2:nrow(ch_df_2)){
  ch_df_2$ch[i] <- 
      case_when(
        is.na(ch_df_2$ch[i]) ~ ch_df_2$ch[i-1], 
        T ~ ch_df_2$ch[i]
      )
}

ch_df_2_wide = 
  ch_df_2 %>% 
  pivot_wider(
    values_from = 'ch',
    names_from = 'group'
  ) %>% na.omit()

ch_df_2_wide$log_diff = with(ch_df_2_wide, log(ch_df_2_wide$`1` / ch_df_2_wide$`2`))

ch_df_2_wide %>% 
  ggplot(aes(x = time, y = log_diff)) + geom_line() + 
  stat_smooth(method = "lm") + 
  stat_smooth(se = T, color = "red")
```

Smooth LOESS red regression line shows that the hazards rate proportion seems to increase until approximate time 75, after which the rates are proportional. There is a lot of variance
in the poprtion of hazards between the two groups.

### (c)

```{r}
cox_3_c <- coxph(Surv(time = time, event= delta) ~ type_descr, data = tongue )

# plot(resid(object = cox_3_c, type="deviance"))

df_base_haz <- basehaz(cox_3_c, centered = FALSE)

ggplot(data = df_base_haz %>% 
         filter(time <= 300), mapping = aes(x = time, y = hazard)) +
    geom_point() +
    labs(x = "Cox-Snell residuals as pseudo observed times",
         y = "Estimated cumulative hazard at pseudo observed times") +
    theme_bw() + theme(legend.key = element_blank()) + 
  stat_smooth(se = T, color = "red") + 
  stat_smooth(method = "lm")
```

I chose to use cox snell residuals to evaluate model fit. 

I removed an outlier that have observed time of around 400. 

Residuals vs time has approximately linear trend, with more 
and more variance occurring at later times. Later observations have a poor fit to the linear trend, which is common for these sets of residuals 

# Exercise 4


### (a)

I am using code below to obtain MLE estimates: 

```{r}
#| echo: true

# https://stackoverflow.com/questions/9113872/interpreting-weibull-parameters-from-survreg

model <- survreg(Surv(time = time, event= delta) ~ 1, data = tongue, dist = "weibull")
summary(model)


shape <- 1/model$scale      
scale <- exp( coef(model) ) 
```

* Shape parameter: `r round(shape, 2)`

* Scale parameter: `r round(scale, 2)`

### (b)

Using output above, we obtain test statistic and a p-value to conduct the test. Null hypothesis: shape parameter is equal to 1. 

* test stat: 27.29

* p-value: 0 

Conclusion: we have enough evidence to conclude that the shape parameter is very different from 1 

### (c)

Plugging in shape and scale MLE parameter into a function, and 
using 50% percentile of weibull distribution, we 
obtain an MLE for median survival time: 

```{r}
#| echo: true 
med.surv <- qweibull(0.5, shape = shape, scale = scale)
med.surv

```

### (d)

```{r}
#| echo: true
model2 <- survreg(Surv(time = time, event= delta) ~ type_descr, data = tongue, dist = "weibull")
summary(model2)

```

* Null hypothesis: no effect of ploidy, i.e. the coefficient is equal to 1 

* Test statistic: - 1.91

* p-value: 0.057

Conclusion: the p-value is above the accepted $\alpha =$ 0.05, but it is extremely close. Therefore, we have a moderately strong belief that the effect of ploidy is statistically significant and is idfferent between the two groups. 
